Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism

### 2024-07-25 index.qmd
section:all; process:write,correct

  Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs of stimuli to determine which one exhibits a higher degree of a specific trait. DCJ has proven effective and reliable across various fields @Pollitt_2012b; @Jones_2015; @vanDaal_et_al_2016; @Bartholomew_et_al_2018; @Lesterhuis_2018; @Bartholomew_et_al_2020; @Marshall_et_al_2020; @Boonen_et_al_2020]. However, despite the method's widespread use, existing literature lacks a clear explanation of the complexities and assumptions underpinning the DCJ system, as well as the plausible mechanisms through which DCJ data are generated. This study addresses these issues by integrating DCJ into the framework of causal inference. Specifically, utilizing a structural approach to causal inference, the study develops a scientific model to clarify the causal assumptions and mechanisms inherent in the DCJ system. It then translates this model into a probabilistic statistical framework to estimate statistical relationships and infer causal connections within the system. This research builds upon Thurstone’s law of comparative judgment [-@Thurstone_1927] and provide a robust probabilistic foundation for the statistical analysis of DCJ data. The findings of this study offer valuable insights for researchers and analysts involved in education and assessment procedures who implement or design DCJ experiments.

@@

<!-- 2. in what has been applied on?, and where is the gap? -->
DCJ has proven effective in assessing competencies and traits predominantly within the educational realm, as demonstrated by @Pollitt_2012b, @Jones_2015, @vanDaal_et_al_2016, @Bartholomew_et_al_2018, @Lesterhuis_2018, @Bartholomew_et_al_2020, and @Marshall_et_al_2020. However, its application transcends education, as exemplified by @Boonen_et_al_2020. The methodology has also evolved to include multiple, as opposed to pairwise comparisons [@Luce_1959; @Placket_1975], and to accommodate comparisons with ordinal outcomes [@Tutz_1986; @Agresti_1992]. Overall, research suggests that DCJ offers an alternative and efficient approach to measurement and evaluation, characterized by its reliability and validity [@Lesterhuis_2018; @vanDaal_2020; @Marshall_et_al_2020]. Nevertheless, despite the method's widespread use, the literature does not offer a clear representation of the plausible mechanisms that generate DCJ data. Particularly, there is no depiction of the complexity and the underlying assumptions of the DCJ system, nor how different assessment factors can potentially influence the observed DCJ outcome.

<!-- 3. What are the factors behind DCJ? -->
According to @Verhavert_et_al_2019 and @vanDaal_2020, several assessment factors interact and influence the method's outcome. These factors include the number and characteristics of the stimuli, their *proximity* in terms of the assessed trait, the number of comparison per stimulus, and the pairing algorithm used. Furthermore, since the method relies on judges' assessments, the number and characteristics of judges, their *discrimination* abilities, and the number of comparisons per judge also play pivotal roles. Moreover, when the stimuli represent sub-units of higher-levels units, factors such as the number and characteristics of these units, along with their *proximity* in terms of the assessed trait, can significantly influence the outcome. For example, @vanDaal_et_al_2016 assessed the academic writing skills of university students (units) using multiple argumentative essays (sub-units).

<!-- 4. Solution to the gap -->
Although several studies have examined the individual impact of these factors on the method's reliability [@Bramley_2015; @Pollitt_2012b; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @vanDaal_et_al_2017; @Gijsen_et_al_2021], none, to the best of the authors' knowledge, have provided a transparent depiction of the DCJ system and the mechanisms generating the DCJ outcome. This study aims to fill this gap by representing DCJ within the causal inference framework. Specifically, using the structural approach to causal inference [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016], the study aims to construct a scientific model. This model will elucidate the underlying assumptions of the DCJ system, providing plausible mechanisms for how the DCJ outcome is generated. Next, using a minimal set of assumptions, the study will translate the scientific model into a probabilistic statistical model. This model will produce statistical estimates to draw inferences about plausible causal relationships within the DCJ system.

<!-- 5. final objective and relevance -->
Ultimately, this research builds upon Thurstone’s law of comparative judgment [-@Thurstone_1927] and provide a robust probabilistic foundation for the statistical analysis of DCJ data. Consequently, the findings of this study offer valuable insights for researchers and analysts involved in education and assessment procedures who implement or design DCJ experiments.

@@

<!-- 4. But which approach are we going to use? -->
However, the structural approach offers a key advantage over the potential outcomes approach: it enables the graphical representation of systems through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020]. In this context, DAGs fulfill the role of a heuristic, effectively conveying the assumed causal structure of a system. They do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAG is accurate [@McElreath_2020].

@@

@@

<!-- 4. But how do we move from targets of inference to estimates -->

@@

@@

@@

>>>>

  Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs of stimuli to determine which one exhibits a higher degree of a specific trait. DCJ has proven effective and reliable across various fields [@Pollitt_2012b; @Jones_2015; @vanDaal_et_al_2016; @Bartholomew_et_al_2018; @Lesterhuis_2018; @Bartholomew_et_al_2020; @Marshall_et_al_2020; @Boonen_et_al_2020]. However, despite the method's widespread use, existing literature lacks a clear explanation of the complexities and assumptions underpinning the DCJ system, as well as the plausible mechanisms through which DCJ data are generated. This study addresses these issues by representing DCJ within the framework of causal inference. Specifically, utilizing a structural approach to causal inference, the study develops a scientific model to clarify the causal assumptions and mechanisms inherent in the DCJ system. It then translates this model into a probabilistic statistical framework to estimate statistical relationships and infer causal connections within the system. This research provides a robust probabilistic foundation for the statistical analysis of DCJ data, building upon Thurstone’s law of comparative judgment [-@Thurstone_1927]. Its findings offer valuable insights for researchers and analysts designing and implementing DCJ experiments.

@@

<!-- 2. in what has been applied on?, and where is the gap? -->
DCJ has proven effective in assessing competencies and traits predominantly within the educational realm, as demonstrated by @Pollitt_2012b, @Jones_2015, @vanDaal_et_al_2016, @Bartholomew_et_al_2018, @Lesterhuis_2018, @Bartholomew_et_al_2020, and @Marshall_et_al_2020. However, its application transcends education, as exemplified by @Boonen_et_al_2020. The methodology has also evolved to include multiple, as opposed to pairwise comparisons [@Luce_1959; @Placket_1975], and to accommodate comparisons with ordinal outcomes [@Tutz_1986; @Agresti_1992]. Overall, research suggests that DCJ offers an alternative and efficient approach to measurement and evaluation, characterized by its reliability and validity [@Lesterhuis_2018; @vanDaal_2020; @Marshall_et_al_2020]. Nevertheless, despite the method's widespread use, existing literature lacks a clear representation of the plausible mechanisms through which DCJ data are generated. Particularly, there is no depiction of the complexity and the assumptions underpinning the DCJ system, nor how different assessment factors can potentially influence the observed DCJ outcome.

<!-- 3. What are the factors behind DCJ? -->
According to @Verhavert_et_al_2019 and @vanDaal_2020, several assessment factors interact and influence the method's outcome. These factors include the number and characteristics of the stimuli, their *proximity* in terms of the assessed trait, the number of comparison per stimulus, and the pairing algorithm used. Furthermore, since the method relies on judges' assessments, the number and characteristics of judges, their *discrimination* abilities, and the number of comparisons per judge also play pivotal roles. Moreover, when the stimuli represent sub-units of higher-levels units, factors such as the number and characteristics of these units, along with their *proximity* in terms of the assessed trait, can significantly influence the outcome. For instance, @vanDaal_et_al_2016 assessed academic writing skills of university students (units) using multiple argumentative essays (sub-units).

<!-- 4. Solution to the gap -->
Although several studies have examined the individual impact of these factors on the method's reliability [@Bramley_2015; @Pollitt_2012b; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @vanDaal_et_al_2017; @Gijsen_et_al_2021], none, to the best of the authors' knowledge, have provided a transparent depiction of the DCJ system and the mechanisms generating the DCJ outcome. This study aims to fill this gap by representing DCJ within the framework of causal inference. Specifically, utilizing a structural approach to causal inference [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016], the study develops a scientific model to clarify the causal assumptions and mechanisms inherent in the DCJ system. Next, using a minimal set of assumptions, the study translates the scientific model into a probabilistic statistical model. This model aims to produce statistical estimates to draw inferences about plausible causal relationships within the DCJ system.

<!-- 5. final objective and relevance -->
Ultimately, this research provides a robust probabilistic foundation for the statistical analysis of DCJ data, building upon Thurstone’s law of comparative judgment [-@Thurstone_1927]. Consequently, its findings offer valuable insights for researchers and analysts designing and implementing DCJ experiments.

@@

<!-- 4. But which approach are we going to use? -->
However, the structural approach offers a key advantage over the potential outcomes approach: it enables the graphical representation of systems through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020]. DAGs are heuristics that effectively convey the assumed causal structure of a system. They do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAG is accurate [@McElreath_2020].

@@

<!-- Graph theory is the branch of mathematics focused on the study of graphs. While graph theory covers a wide array of topics, the structural approach to causal inference incorporates some of its concepts to formally represent causes and counterfactuals. -->

<!-- Graphs are mathematical structures used to model pairwise relations between objects (Gross, Yellen, and Anderson 2018). A graph is a collection of nodes connected by edges. In a directed graph, edges extend from a parent node to a child node, with arrows indicating the direction of the causal influence. In a directed acyclic graph (DAG), the direction of causal influences does not loop back on itself, ensuring that the graph contains no cycles McElreath (2020). -->

<!-- To draw a DAG use the exclusion restriction assumption: a variable should not appear in the graph if it is not a part of the causal mechanisms that generate the outcome under analysis (see Neal, 2020). -->

@@

<!-- 4. But how do we move from targets of inference to estimates -->
<!-- But how can we move from heuristics and vague mathematic notation, to probabilistic causal analysis. -->

@@

<!-- From Lawson (2015), another way to see DCJ experiments I'd that you are grouping heterogeneous experimental units (objects) into homogeneous ones (subjects), which are your "blocks" -->

@@

<!-- It is more likely that we can decide the characteristics of the objects we want to include in the study (independent variables), ensuring groups that are more balanced. However, in the case of subjects sometimes the independent variables cannot be assigned (e.g., hearing status) -->

@@

<!-- "An appropriate analysis of data cannot be completed without knowledge of what experimental design was used and how the data was collected, and conclusions are not reliable if they are not justified by the proper modeling and analysis of the data" (Lawson, 2015, p. 7) -->

<!-- "In many cases where observational units are different than the experimental units, the observational units will not be independent" (p. 177) -->
